%% The first command in your LaTeX source must be the \documentclass command.
%%
%% Options:
%% twocolumn : Two column layout.
%% hf: enable header and footer.
\documentclass[
  % twocolumn,
  % hf,
]{ceurart}

%%
%% One can fix some overfulls
\sloppy

%%
%% Minted listings support
%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
\usepackage{listings}
%% auto break lines
\lstset{breaklines=true}

\usepackage{rotating}
\usepackage{hyperref}
\usepackage{longtable}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% Rights management information.
%% CC-BY is default license.
\copyrightyear{2025}
\copyrightclause{Copyright for this paper by its authors.
  Use permitted under Creative Commons License Attribution 4.0
International (CC BY 4.0).}

%%
%% This command is for the conference information
\conference{Workshop on Computational Drama Analysis, 3 September
2025, Berlin, DE}

%%
%% The "title" command
\title{The Measure of a Messenger: Meter and lexis as potential avenues
of differentiation}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
\author[1]{Charles Pletcher}[%
  orcid=0000-0003-2734-5511,
  email=charles.pletcher@tufts.edu,
  url=https://github.com/pletcher,
]
\cormark[1]
\address[1]{Tufts University, Department of Classical Studies, 10
Upper Campus Road, Medford, Massachusetts 02155, United States}

%% Footnotes
\cortext[1]{Corresponding author.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  This paper argues that messenger speeches (\textit{Botenberichte})
  in the ancient tragedies of Aeschylus, Sophocles, and Euripides
  have not received the computational attention that they deserve. I
  first draw attention to this gap by surveying traditional literary
  approaches to the genre, showing how the definition of ``messenger
  speech'' is already problematic from a qualitative perspective.
  Next, I present a series of initial inquiries into quantitative and
  computational approaches to messenger speeches, beginning with a
  stylometric comparison to Homeric epic and proceeding through topic
  modeling and, using the editions prepared by DraCor, network
  analysis. In conclusion, I offer a brief discussion of these
  preliminary findings and potential avenues for further inquiry.
\end{abstract}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\begin{keywords}
  tragedy \sep
  messenger speeches \sep
  Homer \sep
  topic modeling \sep
  network theory
\end{keywords}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}

In the first pages of \textit{Narrative in Drama: The Art of the Euripidean
Messenger-Speech}, Irene J. F. de Jong offers the following definition of
(Euripidean) messenger-speeches: ``The Euripidean messenger-speech or ἀγγελικὴ
ῥῆσις is a long, continuous speech in which a messenger reports events that
have taken place off-stage. The content of this speech is a narrative, more
particularly a first-person narrative…'' \cite[1–2]{deJong1991}. She goes on to
argue that Euripides makes signficantly greater use of first-person narrative
than Sophocles and Aeschylus, citing \cite[2 n. 4]{deJong1991}

\begin{itemize}
  \item Aeshylus, \textit{Persians} 480–514
  \item Aeschylus, \textit{Agamemnon} 636–80
  \item Sophocles, \textit{Antigone} 407–40, 1192–1243
  \item Sophocles, \textit{Oedipus Tyrannus} 1237–85
  \item Sophocles, \textit{Oedipus Colonus} 1586–1666
\end{itemize}

As the only examples of first-person narrative outside of Euripides' oeuvre.
Although de Jong takes great pains (and around sixty pages) establishing the
first-person focalization of Euripides' messenger-speeches, a cursory glance
at, for example, the Tutor's speech in Sophocles' \textit{Electra} (680–763)
casts enough doubt on de Jong's criteria that it raises further questions about
not only Euripidean messenger-speech exceptionalism but messenger-speech as a
category in its own right. \footnote{The Tutor's speech is rightly considered
  an exemplary ``messenger-type speech'' in its own right (so
  \cite{Marshall2006} and \cite{Kells1973}, among others), and it is largely
  for this reason that its apparent violation of Euripidean messenger-speech
exceptionalism raises other questions.} In this paper, I present three
computational null results that cast further doubt on messengers'
exceptionalism. By showing how ordinary messengers' speeches are, I mean both
to highlight their dramatic---as opposed to narrative---place and to
demonstrate how their performances contribute to tragedy's dramatic
negotiations for truth.\footnote{I use ``messenger-speech'' when referring to
  de Jong's concept, and ``messenger's speech,'' ``messengers' speeches,'' or
  ``messenger-type speech'' when referring to the general category of dramatic
monologue.}

The paper begins with a brief and necessarily incomplete overview of the
massive (and growing) bibliography on messengers and their speeches. I then
discuss the methodologies and experiments by which this paper's results were
obtained, beginning with the library that was used for scanning iambic
trimeters, the spoken meter of Greek tragedy and thus the meter for most
messenger-type (Euripides singing Phrygian in \textit{Orestes} marking a
notable exception). The first two experiments examine two interrelated ways in
which meter might be supposed to distinguish messengers' speeches from other
characters' speech, but which instead show that meter does not sufficiently
distinguish messengers from other tragic speakers. The third experiment uses
TF--IDF and 4-character-grams to cluster the approximately 17,000 lines of
iambic trimeter in tragedy with all of the speeches from the Homeric corpus,
obtained through the Digital Initiative for Classics: Epic Speeches (DICES)
API. These experiments and their results are far from exhaustive, but they
offer a new way of understanding the place of messengers in drama---a place,
moreover, that appears no less ``dramatic'' than their traditionally less
``narrative'' peers.

\section{A brief history of \textit{Botenberichte}}

The usual history of (so-called) ``messenger speeches'' in drama begins with
Homer: the ancient bard's narrative \textit{enargeia}, to borrow a term from
Pseudo-Longinus, is said to animate the off-stage events that news-bringers
narrate. Irene J. F. de Jong has perhaps most famously applied a narratological
lens to both Homer and tragic messengers, followed a few years later by Barbara
Goward and James Barrett with their own seminal works on messenger speeches in
2002 \cite{deJong2011, deJong1991, Goward1999, Barrett2002}. De Jong championed
the view that messengers are not featureless conduits for the dramatist's
voice but are instead focalizers with bias and selective memory. James
Barrett then qualified this view by noting that messengers' narrative
effectiveness is inversely proportional to their stage presence: ``The
messenger's own body,'' he writes, ``is implicitly—and explicitly, on
occasion—a hindrance, inasmuch as his success as a messenger depends upon
his acquiring the invulnerability granted by freedom from bodily
constraints'' \cite[100]{Barrett2002}. More recently, Margaret Dickin
showed how the roles of messengers in tragedy increased over time and
became ``vehicles for star-power,'' while others, like Felix Budelmann and
Evert van Emde Boas, explored how messengers direct the audience's
affective energies \cite{Dickin2009, Budelmann.vanEmdeBoas2020}. Florence
Yoon's interventions have further disrupted the ``category'' of messengers
by revealing the mistakes of painting with too broad a brush:
\textit{angeloi}, messengers proper, and \textit{kerykes}, ``heralds,''
perform different functions in tragedy \cite{Yoon2022}.

The chief difficulty running through much of the literature on messenger
speeches has to do with defining who, precisely, is a news-bringer\footnote
{Along with ἄγγελοι, ``messengers,'' I also count κήρυκες, ``heralds,'' and
various other unnamed deliverers of information. For the purposes of the
computational studies that follow, however, I divide these speakers into
different classes, described below. As I hope to show, any metrical
distinctions between classes of characters in tragedy are artificial at
best.} and what, precisely, is a messenger speech. Part of the difficulty
arises from the fact that named characters at times appear to deliver
speeches that resemble messenger speeches: nearly every named character in
Sophocles'
\textit{Trachiniae} does so, as Sir Richard Jebb noted over a century ago
\cite{Jebb1902}; and even Clytaemestra's famous "Beacon Speech" in Aeschylus'
\textit{Agamemnon} has resembles speeches of news-bringers in its
subject-matter. Thus, some scholars insist on anonymity as a prime criterion
for the deliverer of a messenger speech, but such an insistance runs the risk
of obscuring the similarities that so-called ``messenger speeches'' have with
respect to other monologues in drama. These similarities matter, I argue,
because they highlight the various strategies that drama uses to interrogate
the fraught and problematic delivery of information that the recipients cannot
verify. The tragedians, to state the matter bluntly, offer some of the earliest
extended encounters with ``fake news.''

Yet despite the steady stream of innovative approaches to messenger speeches in
tragedy, there have been few attempts to address the various classes of
news-bringers from a computational angle. Where such examinations have
occurred, they have, to my knowledge, taken place in the context of broader
computational studies of drama, such as the work of Julia Jennifer Beine, Frank
Fischer, and Viktor Illmer presented at the Digital Humanities Conference in
2024; the examination of the chorus by María Teresa Santa María Fernández and
Monika Dabrowska; or the early work on social network theory and Greek tragedy
by Jeff Rydberg-Cox \cite{Beine.etal2024, Fernandez.Dabrowska2023,
Rydberg-Cox2011}.

In this paper, I undertake three quantitative approaches to meter in tragedy,
with a focus on whether messengers and their ilk exhibit a discernible
metrical signature by which they might be differentiated from other speakers.
I investigate three specific phenomena: (1) \textbf{the number of resolutions
per line of iambic trimeter}, that is, the number of times that a single long
syllable in an iamb is replaced by two short syllables; (2) \textbf
{the placement of resolutions in characters' trimeters}, which has not, to my
knowledge, been used as a way to distinguish between classes of speakers,
although the \textit{frequency} of resolution has been used for dating the
plays of Euripides and Sophocles; and (3) a corollary of resolutions, \textbf
{the number of dactyls per line of a character's trimeter}, by way of
exploring the hypothesis that such ``dactylic events'' lend a line more epic
character.

At the risk of discouraging the reader from continuing, all of the experiments
turn up null results, with extremely weak correlation co-efficients at best:
the analyzed features are insufficient to distinguish messengers from other
classes of tragic characters. Yet the reasons for this lack of differentation
can still lead to a more nuanced understanding of the \textit{dramatic} place
of messengers in tragedy, as opposed to their relegation to ``\textit
{narrative} in drama.'' By exploring the ways that we can measure a
character's ``messenger-type-ness,'' or the extent to which they perform a
``messenger function'' (to borrow a term from C. W. Marshall \cite
{Marshall2006}), we shed new light on this often-misunderstood class of
dramatic actors and thus better understand how they reflect their culture's
relationship to secondhand information.

\section{Data preparation}

Data were prepared using TEI XML files from DraCor and the Perseus Project.
For determining scansion, a rules-based library was built in the Elixir
programming language. Although the deviation from Python might seem unusual,
Elixir has the benefit of platform stability---anecdotally, Python projects
break approximately every six months, while I have been able to maintain
Elixir-based experiments for nearly a decade without major interruption--- as
well as first-class support for multi-threaded and parallel computing,
techniques that will be beneficial for larger corpora. This early
programming-language decision will mean straightforward parallelization as/if
the corpus grows. It also means that these tools are reusable for larger
projects, even if the developers of those projects are not familiar with
Elixir: they need only install the Elixir runtime and run the pre-written
processing scripts to output their own scansion analysis.

Experiments were run in Livebook, an Elixir-based corollary to Jupyter
notebooks. The relevant Livebooks and supporting files are available upon
request. (They are not currently hosted on GitHub and so cannot be anonymized.)

For the sake of efficiency, the rules-based scansion parser was built with the
assistance of Claude Code. The author verified the scansion results, and the
author wrote all of the experiments and conducted the analysis without AI
assistance.

The scansion parser first syllabifies a text and then parses line by line to
determine whether the line can scan as valid iambic trimeter. Out of 44,482
lines of tragedy in the corpus, 17,341 scanned as valid trimeters;
81 were skipped due to parsing errors. The corpus of iambic trimeters
represents less than 40\% of the lines of tragedy in the corpus, which
restricts the kinds of experiments that will yield meaningful results.

For all of the experiments, each speaker was labeled with one of eight
``speaker classes'': (1) messenger, (2) herald, (3) chorus, (4) royalty,
(5) divinity, (6) slave, (7) prophet, or (8) other. These labels were first
applied automatically to unambiguous characters---all Ἄγγελοι were labeled
``messenger,'' for example. Subsequently, a list of ambiguous characters
was generated and labeled by hand.

\section{Parsing iambic trimeter}

The rules-based parser works as follows (visualized also in the diagram
below): (1) First, the raw Greek text is read to determine whether it is
Beta Code or polytonic Greek. These experiments work exclusive with
polytonic Greek, so this step is essentially a noop. (2) Second, the raw text is  tokenized by grapheme (letters with diacritical
markings attached). (3) Subsequently, the graphemes are grouped by syllables, accounting
for diphthongs, elision, and double consonants. Compounds such as εἰσ-φέρω are not
explicitly handled, as doing so would add out-of-scope morphological analysis
to this scanner with minimal improvement on the scansion analysis: for example, in the case
of εἰσ-φέρω, εἰ- will always scan long, and the epsilon in φέρω will always scan
short, whether the compound is grouped correctly as εἰσ-φέρω or incorrectly as
εἰ-σφέρω. (4) Syllables are classified as short, long, anceps, or unknown according
to standard rules of Ancient Greek scansion. According to the usual convention
in tragic verse, \textit{muta cum liquida} (a stop consonant followed by a lambda or a rho)
does not result in the usual lengthening that applies to other short vowels followed
by two consonants. Other double-consonant clusters will cause preceding short vowels to be scanned long by this scanner.\footnote{Unknown vowel quantities are logged in a dictionary for later manual annotation.
This annotation process has not yet been completed.} (5) The scanned syllables are mapped
onto the iambic trimeter schema of three dipodies (in its most basic form: x — u — | x — u — | x — u —). (6) Features such as resolution and dactylic events are labeled.

\begin{figure}
\caption{Flow chart of the parsing and scanning process}
\begin{center}
\begin{tabular}{c}
(1) raw text \\
\downarrow{} \\
(2) grapheme tokenization \\
\downarrow{} \\
(3) syllabification \\
\downarrow{} \\
(4) quantification (labeling of short, long, and anceps syllables) \\
\downarrow{} \\
(5) parsing (mapping to iambic trimeter) \\
\downarrow{} \\
(6) feature extraction (label resolutions and dactylic events)
\end{tabular}
\end{center}
\end{figure}

Dactylic events refer specifically to the possibility of a long anceps followed by a resolved
iambic ``up-beat,'' i.e., the familiar epic — u u (long short short). A hypothesis concerning the ``epic-ness'' of messengers in tragedy might suppose
that they make greater use of dactyls than other classes of speakers. This hypothesis
turns out false, however.

Due to the incompleteness of the ambiguous length dictionary (see footnote \textit{supra}),
the 17,341 lines obtained for this analysis is almost certainly an undercount. However,
it should still be sufficiently many lines---40\% of extant tragedy---to detect patterns
by character class.

\section{Resolution count}

We begin by counting resolutions in iambic trimeters. The frequency of
resolution offers a sense of how often each character class adds more
syllables than are strictly necessary to a line of iambic trimeter. The
strength of the relationship between speaker class and resolution count was
tested using [Harald] Cramér's $φ_c$ (also known as Cramér's V). As Cramér
writes, ``The correlation coefficient and the correlation ratio both server
to characterize, in the sense explained above, the `degree of dependence'
between two variables.... $φ^2 = 0$ when and only when the variables are
independent.'' \cite[282]{Cramer2016} Cramér goes on to explain that $φ_c$
will range between \(0\) (no correlation) and \(1\)(full corrleation). Thus,
by measuring Cramér's $φ_c$ for speaker class and number of resolutions, we
can determine whether the two are related and the strength of that relation.

For 8 speaker classes (7 degrees of freedom), the correlation with resolution
count was 0.107---meaning a small but significant connection between the two
variables. The figure below shows the relative frequency of resolutions per
total lines by speaker class.

\begin{figure}
\caption{Absolute and relative frequencies by speaker class for total resolutions}
\begin{center}
\begin{tabular}{c | c | c | c}
\textbf{speaker class} & \textbf{resolutions} & \textbf{total lines} & \textbf{rel. freq. resolved} \\
prophet & 60 & 445 & 0.134 \\
messenger & 283 & 2795 & 0.101 \\
other & 204 & 2210 & 0.092 \\
royalty & 1754 & 23562 & 0.074 \\
divine & 194 & 2652 & 0.073 \\
slave & 60 & 1422 & 0.042 \\
herald & 28 & 665 & 0.042 \\
chorus & 246 & 10731 & 0.023
\end{tabular}
\end{center}
\end{figure}


\section{Resolution placement}


\begin{figure}
\caption{Cramér's $φ_c$ for each potential resolution in iambic trimeter and its association with speakers' character class}
\begin{itemize}
  \item Cramér's $φ_c$ (speaker class and resolution at position 2): 0.0479
  \item Cramér's $φ_c$ (speaker class and resolution at position 4): 0.0421
  \item Cramér's $φ_c$ (speaker class and resolution at position 6): 0.0749
  \item Cramér's $φ_c$ (speaker class and resolution at position 8): 0.0468
  \item Cramér's $φ_c$ (speaker class and resolution at position 10): 0.0154
\end{itemize}
\end{figure}

Although these correlations appear to be small, given the high number of degrees
of freedom (7, for 8 speaker classes) in each calculation, these represent
small but meaningful correlations for most resolution placements and character
class.

Indeed, we can see small but meaningful differences in the absolute and relative
frequencies for each resolution for every position but 10 (omitted for brevity).


\begin{figure}
\caption{Absolute and relative frequencies by speaker class for resolutions in position 2}
\begin{center}
\begin{tabular}{c | c | c | c}
\textbf{speaker class} & \textbf{resolutions} & \textbf{total lines} & \textbf{rel. freq. resolved} \\
prophet & 14 & 445 & 0.0314 \\
other & 47 & 2210 & 0.0212 \\
messenger & 44 & 2795 & 0.0157 \\
royalty & 354 & 23562 & 0.0150 \\
divine & 39 & 2652 & 0.01470 \\
slave & 10 & 1422 & 0.007 \\
chorus & 52 & 10731 & 0.004 \\
herald & 1 & 665 & 0.002
\end{tabular}
\end{center}
\end{figure}

\begin{figure}
\caption{Absolute and relative frequencies by speaker class for resolutions in position 4}
\begin{center}
\begin{tabular}{c | c | c | c}
\textbf{speaker class} & \textbf{resolutions} & \textbf{total lines} & \textbf{rel. freq. resolved} \\
prophet & 14 & 445 & 0.0314 \\
messenger & 47 & 2795 & 0.0168 \\
other & 32 & 2210 & 0.0145 \\
divine & 33 & 2652 & 0.0124 \\
royalty & 273 & 23562 & 0.012 \\
slave & 13 & 1422 & 0.009 \\
chorus & 49 & 10731 & 0.005 \\
herald & 1 & 665 & 0.002
\end{tabular}
\end{center}
\end{figure}


\begin{figure}
\caption{Absolute and relative frequencies by speaker class for resolutions in position 6}
\begin{center}
\begin{tabular}{c | c | c | c}
\textbf{speaker class} & \textbf{resolutions} & \textbf{total lines} & \textbf{rel. freq. resolved} \\
prophet & 21 & 445 & 0.0471 \\
messenger & 130 & 2795 & 0.0465 \\
divine & 96 & 2652 & 0.0362 \\
other & 79 & 2210 & 0.0357 \\
royalty & 787 & 23562 & 0.033 \\
herald & 18 & 665 & 0.0270 \\
slave & 29 & 1422 & 0.0204 \\
chorus & 79 & 10731 & 0.007
\end{tabular}
\end{center}
\end{figure}


\begin{figure}
\caption{Absolute and relative frequencies by speaker class for resolutions in position 8}
\begin{center}
\begin{tabular}{c | c | c | c}
\textbf{speaker class} & \textbf{resolutions} & \textbf{total lines} & \textbf{rel. freq. resolved} \\
prophet & 9 & 445 & 0.020 \\
messenger & 56 & 2795 & 0.020 \\
other & 38 & 2210 & 0.017 \\
royalty & 295 & 23562 & 0.013 \\
herald & 7 & 665 & 0.011 \\
divine & 24 & 2652 & 0.009 \\
slave & 8 & 1422 & 0.006 \\
chorus & 39 & 10731 & 0.004
\end{tabular}
\end{center}
\end{figure}


As expected by Cramér's $φ_c$, position 6 has the strongest correlation with speaker class.
In all positions, we see that prophets and messengers tend to have the highest relative
frequency of resolutions, suggesting a revelatory valence on messengers' speeches that
resembles the prophecies of Tiresias---even though Tiresias makes predictions while
messengers refer to past events.

\section{Dactylic events}









\section{Declaration on Generative AI}

The author built the metrical parsing library with the
help of Anthropic's Claude Code. No AI was used in the calculations
or analysis that ultimately make up the bulk of this paper.

\bibliography{zotero-refs}

\end{document}
